---
title: "Spatial correlation modeling of environmental extremes with HazardMaps"
author: "Maria Valentina Clavijo Mesa"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Spatial correlation modeling of environmental extremes with HazardMaps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  fig.align  = "center",
  message    = FALSE,
  warning    = FALSE
)
```

<style>
.note-box {
  background-color: #EAF4FB;
  border-left: 4px solid #2980B9;
  padding: 12px 16px;
  margin: 16px 0;
  border-radius: 4px;
}
.result-box {
  background-color: #EAFAF1;
  border-left: 4px solid #27AE60;
  padding: 12px 16px;
  margin: 16px 0;
  border-radius: 4px;
}
.warn-box {
  background-color: #FEF9E7;
  border-left: 4px solid #F39C12;
  padding: 12px 16px;
  margin: 16px 0;
  border-radius: 4px;
}
</style>

## Overview

**HazardMaps** provides a unified workflow for spatial dependence modeling
of environmental extremes — from raw gridded datasets to spatially coherent
pseudoscenarios. The package supports regular longitude-latitude grids
(e.g. ERA5, Copernicus) and rotated curvilinear grids (e.g. EURO-CORDEX)
without any change in the user interface.

<div class="note-box">
**Example dataset.** This vignette uses daily maximum wind-gust speed over
Northeast Italy as a running example. The workflow applies unchanged to any
gridded environmental variable: precipitation, temperature, wave height, etc.
The package was developed as part of the research presented in
Clavijo Mesa et al. (2026a, 2026b), whose datasets are included as built-in
example files. Users who need to prepare raw Copernicus or EURO-CORDEX
downloads before using the package can find helper R scripts in
`inst/scripts/`, accessible via
`system.file("scripts", package = "HazardMaps")`.
</div>

Two datasets are used throughout:

- **Copernicus** — regular longitude-latitude grid, ERA5-based, 195 cells,
  1994–2021
- **EURO-CORDEX** — rotated curvilinear grid, RCM simulation, 1872 cells
  (1568 land), 1970–2005

The workflow has seven steps:

1. Ingest and prepare data
2. Select extreme-event days
3. Fit marginal distributions
4. Estimate spatial dependence
5. Fit copula and stochastic field models
6. Simulate pseudoscenarios
7. Validate with cluster-size statistics

```{r load, message = FALSE}
library(HazardMaps)
```

---

## Step 1 — Ingest and prepare data

The starting point is a CF-convention NetCDF file. `hm_read_netcdf()` reads
it and detects the grid type automatically — no configuration needed.

```{r ingest}
wind_cop <- hm_read_netcdf(
  system.file("extdata", "Copernicus_dailymax_1994_2021.nc",
              package = "HazardMaps"),
  var = "i10fg"
)

wind_euc <- hm_read_netcdf(
  system.file("extdata", "EuroCordex_wsgsmax_1970_2005.nc",
              package = "HazardMaps"),
  var = "wsgsmax"
)

wind_cop$meta$grid_type
wind_euc$meta$grid_type
```

`hm_standardize_coords()` flattens the coordinate arrays into a single
data.frame with one row per grid cell — the format all downstream functions
expect. `hm_decode_time()` converts CF-style time values (e.g.
*"days since 1970-01-01"*) to R `Date` objects.

```{r standardize}
wind_cop <- hm_standardize_coords(wind_cop)
wind_cop <- hm_decode_time(wind_cop)
wind_cop <- hm_get_grid_geometry(wind_cop)

wind_euc <- hm_standardize_coords(wind_euc)
wind_euc <- hm_decode_time(wind_euc)
wind_euc <- hm_get_grid_geometry(wind_euc)

cat("Copernicus: ", nrow(wind_cop$coords), "cells |",
    length(wind_cop$time), "days\n")
cat("EURO-CORDEX:", nrow(wind_euc$coords), "cells |",
    length(wind_euc$time), "days\n")
```

`hm_plot_spatial_domain()` visualises the study area. It draws the correct
cell geometry automatically — straight lines for regular grids, curvilinear
lines for rotated grids.

```{r domain-cop, fig.cap = "Copernicus regular grid domain."}
hm_plot_spatial_domain(wind_cop, dataset = "copernicus",
                       title = "Copernicus domain", show_map = TRUE)
```

```{r domain-euc, fig.cap = "EURO-CORDEX rotated grid domain."}
hm_plot_spatial_domain(wind_euc, dataset = "eurocordex",
                       title = "EURO-CORDEX domain", show_map = TRUE)
```

---

## Step 2 — Select extreme-event days

Only days with at least one grid cell at or above a threshold are kept.
Everything below the threshold is noise for dependence modeling purposes.

```{r extreme-events}
wind_cop_ext <- hm_select_extreme_events(wind_cop, threshold = 25)
wind_euc_ext <- hm_select_extreme_events(wind_euc, threshold = 25)

cat("Copernicus: ", wind_cop_ext$meta$filter$n_days_retained,
    "extreme-event days out of", wind_cop_ext$meta$filter$n_days_original, "\n")
cat("EURO-CORDEX:", wind_euc_ext$meta$filter$n_days_retained,
    "extreme-event days out of", wind_euc_ext$meta$filter$n_days_original, "\n")
```

`hm_plot_exceedance()` maps how often each cell exceeded the threshold over
the full period.

```{r exceedance-cop, fig.cap = "Exceedance count — Copernicus."}
hm_plot_exceedance(wind_cop, threshold = 25, metric = "count",
                   dataset = "copernicus", show_map = TRUE,
                   title = "Wind-gust exceedance count — Copernicus")
```

```{r exceedance-euc, fig.cap = "Exceedance count — EURO-CORDEX."}
hm_plot_exceedance(wind_euc, threshold = 25, metric = "count",
                   dataset = "eurocordex", show_map = TRUE,
                   title = "Wind-gust exceedance count — EURO-CORDEX")
```

`hm_plot_exceedance_time()` shows whether the frequency of extreme events
changes over the observation period.

```{r time-cop, fig.cap = "Extreme-event days per year — Copernicus."}
hm_plot_exceedance_time(wind_cop, threshold = 25, by = "year",
                        title = "Extreme-event days per year — Copernicus")
```

```{r time-euc, fig.cap = "Extreme-event days per decade — EURO-CORDEX."}
hm_plot_exceedance_time(wind_euc, threshold = 25, by = "decade",
                        title = "Extreme-event days per decade — EURO-CORDEX")
```

---

## Step 3 — Fit marginal distributions

Before estimating spatial dependence, each cell needs its own marginal
distribution. This separates the marginal behavior from the dependence
structure — a standard approach in multivariate extreme-value analysis.

`hm_to_matrix()` reshapes the data into a time-by-cell matrix.
`hm_filter_land_cells()` removes columns with no valid data — mandatory for
EURO-CORDEX, whose domain includes sea cells.

```{r matrix}
obs_cop <- hm_to_matrix(wind_cop_ext)
obs_euc <- hm_to_matrix(wind_euc_ext)

obs_cop <- hm_filter_land_cells(obs_cop)
obs_euc <- hm_filter_land_cells(obs_euc)

cat("Copernicus matrix :", nrow(obs_cop), "days x", ncol(obs_cop), "cells\n")
cat("EURO-CORDEX matrix:", nrow(obs_euc), "days x", ncol(obs_euc), "cells\n")
```

`hm_fit_marginals()` fits six candidate families to each cell by maximum
likelihood and selects the best by BIC. The six families are: Weibull,
Gamma, Lognormal, Gumbel, Zero-truncated Gaussian, and Zero-truncated
Laplace.

```{r marginals, cache = TRUE}
marginals_cop <- hm_fit_marginals(obs_cop, criterion = "bic")
marginals_euc <- hm_fit_marginals(obs_euc, criterion = "bic")
```

```{r marginals-summary}
cat("Copernicus — best distribution per cell:\n")
print(sort(table(marginals_cop$best_dist), decreasing = TRUE))

cat("\nEURO-CORDEX — best distribution per cell:\n")
print(sort(table(marginals_euc$best_dist), decreasing = TRUE))
```

<div class="result-box">
**What to look for.** Weibull dominates in both datasets — this is expected
for wind-gust extremes. The spatial map below shows which cells deviate from
Weibull and why: coastal cells and Alpine valleys tend to favour Gamma or
Zero-truncated Gaussian due to their different tail behavior.
</div>

```{r best-dist-cop, fig.cap = "Best-fitting distribution at each cell — Copernicus."}
hm_plot_best_dist(wind_cop_ext, marginals_cop,
                  title = "Best-fitting marginal — Copernicus",
                  show_map = TRUE)
```

```{r best-dist-euc, fig.cap = "Best-fitting distribution at each cell — EURO-CORDEX."}
wind_euc_land        <- wind_euc_ext
wind_euc_land$coords <- attr(obs_euc, "coords")

hm_plot_best_dist(wind_euc_land, marginals_euc,
                  title = "Best-fitting marginal — EURO-CORDEX",
                  show_map = TRUE)
```

```{r dist-freq, fig.cap = "Frequency of best-fitting families across all cells."}
hm_plot_dist_frequency(marginals_cop,
                       title = "Distribution frequency — Copernicus")
hm_plot_dist_frequency(marginals_euc,
                       title = "Distribution frequency — EURO-CORDEX")
```

The Probability Integral Transform (PIT) converts each observation to the
uniform scale using its fitted CDF, then to Gaussian scores via the probit
transform. The Gaussian scores (`Z`) feed the spatial dependence estimation;
the uniform values (`U`) feed the copula.

```{r pit}
pit_cop <- hm_pit_transform(obs_cop, marginals_cop)
pit_euc <- hm_pit_transform(obs_euc, marginals_euc)

cat("Copernicus — cells transformed:", pit_cop$n_cells_ok,
    "| skipped:", pit_cop$n_cells_skipped, "\n")
cat("EURO-CORDEX — cells transformed:", pit_euc$n_cells_ok,
    "| skipped:", pit_euc$n_cells_skipped, "\n")
```

---

## Step 4 — Estimate spatial dependence

The empirical spatial correlation is measured with Kendall's τ, which is
rank-based and does not assume any distribution for the data.

`hm_distance_matrix()` computes the great-circle distance in kilometres
between every pair of cells. `hm_empirical_kendall()` estimates the
pairwise Kendall correlation from the Gaussian scores.

<div class="warn-box">
**Computational note.** For large grids (> 500 cells), computing all N(N-1)/2
pairs is expensive. Use `cell_idx` to pass a representative subsample — as
done here for EURO-CORDEX.
</div>

```{r kendall, cache = TRUE}
distances_cop <- hm_distance_matrix(attr(obs_cop, "coords"))
kendall_cop   <- hm_empirical_kendall(pit_cop$Z)

land_idx      <- which(!is.na(marginals_euc$best_dist))[1:50]
distances_euc <- hm_distance_matrix(attr(obs_euc, "coords")[land_idx, ])
kendall_euc   <- hm_empirical_kendall(pit_euc$Z, cell_idx = land_idx)
```

`hm_empirical_correlogram()` bins the pairwise correlations by distance to
produce a smooth summary of how correlation decays with separation.

```{r correlogram}
corg_cop <- hm_empirical_correlogram(kendall_cop, distances_cop, n_bins = 20)
corg_euc <- hm_empirical_correlogram(kendall_euc, distances_euc, n_bins = 15)
```

```{r correlogram-plot, fig.cap = "Empirical spatial correlogram for both datasets."}
hm_plot_correlogram(corg_cop,
                    title = "Empirical correlogram — Copernicus")
hm_plot_correlogram(corg_euc,
                    title = "Empirical correlogram — EURO-CORDEX (50-cell subsample)")
```

<div class="result-box">
**Interpretation.** Correlation is highest at short distances (~0.75–0.80)
and decreases steadily with increasing separation, reaching near-zero beyond
250–300 km. This pattern confirms strong local spatial dependence and weak
long-range dependence — consistent with synoptic-scale windstorm events
that affect areas of 50–200 km.
</div>

---

## Step 5 — Fit dependence models

Two model families are available and compared.

### Copula models

Vine copulas decompose the joint distribution of all cells into a cascade
of bivariate copulas. Three structures are available: **C-vine** (one
central node drives all others), **D-vine** (a chain structure), and
**Gaussian** (elliptical baseline).

```{r copulas, cache = TRUE}
uniform_small   <- pit_cop$U[, 1:30]
marginals_small <- marginals_cop[1:30, ]

copula_cvine <- hm_fit_copula(uniform_small, type = "cvine")
copula_dvine <- hm_fit_copula(uniform_small, type = "dvine")
copula_gauss <- hm_fit_copula(uniform_small, type = "gaussian")

cat("Cells:", copula_cvine$n_cells,
    "| Observations:", copula_cvine$n_obs, "\n")
```

### Stochastic field model

The stochastic field approach fits a parametric covariance function directly
to the empirical correlogram, then generates realisations via the
Karhunen-Loève Expansion (KLE). Four covariance families are compared
automatically: **Matérn**, **Exponential**, **Gaussian**, and **Spherical**.

```{r sf-fit}
sf_model_cop <- hm_fit_correlation_model(corg_cop)
sf_model_euc <- hm_fit_correlation_model(corg_euc)

cat("Copernicus  — best:", sf_model_cop$best_family,
    "| RMSE:", round(sf_model_cop$best_rmse, 4), "\n")
cat("EURO-CORDEX — best:", sf_model_euc$best_family,
    "| RMSE:", round(sf_model_euc$best_rmse, 4), "\n")

cat("\nFamily comparison — Copernicus:\n")
print(sf_model_cop$all_fits)

cat("\nFitted parameters:\n")
print(round(sf_model_cop$best_params, 4))
```

<div class="result-box">
**How the best family is chosen.** All four covariance families are fitted
by weighted least squares to the empirical correlogram. The one with the
lowest RMSE is selected automatically and stored in `model$best_family`.
No manual choice is needed.
</div>

---

## Step 6 — Simulate pseudoscenarios

Each model generates a matrix of synthetic extreme-event fields. Every row
is one spatially coherent scenario — a plausible simultaneous wind-gust
intensity at all grid cells.

```{r sim-copula}
scenarios_cvine <- hm_simulate_copula(copula_cvine, marginals_small,
                                       n_scenarios = 500L, seed = 42L)
scenarios_dvine <- hm_simulate_copula(copula_dvine, marginals_small,
                                       n_scenarios = 500L, seed = 42L)
scenarios_gauss <- hm_simulate_copula(copula_gauss, marginals_small,
                                       n_scenarios = 500L, seed = 42L)

cat("Copula scenarios — dimensions:", dim(scenarios_cvine)[1], "x",
    dim(scenarios_cvine)[2], "\n")
cat("Value range: [",
    round(min(scenarios_cvine, na.rm = TRUE), 2), ",",
    round(max(scenarios_cvine, na.rm = TRUE), 2), "] m/s\n")
```

```{r sim-sf, cache = TRUE}
scenarios_sf <- hm_simulate_sf_kle(
  sf_model_cop, distances_cop, marginals_cop,
  n_scenarios = 500L, var_retain = 0.95, seed = 42L
)

cat("SF scenarios — dimensions:", dim(scenarios_sf)[1], "x",
    dim(scenarios_sf)[2], "\n")
cat("KLE modes retained:", attr(scenarios_sf, "k_modes"),
    "| Variance explained:", round(attr(scenarios_sf, "var_explained"), 3), "\n")
cat("Value range: [",
    round(min(scenarios_sf, na.rm = TRUE), 2), ",",
    round(max(scenarios_sf, na.rm = TRUE), 2), "] m/s\n")
```

---

## Step 7 — Validate with cluster-size statistics

A good dependence model should reproduce the **spatial footprint** of
observed events — not just the marginal distribution of each cell in
isolation.

`hm_cluster_size()` counts, for each day, how many cells simultaneously
exceeded the threshold. This gives a simple but powerful summary of the
spatial extent of each event.

```{r cluster-obs}
cluster_obs <- hm_cluster_size(obs_cop, threshold = 25,
                                time = attr(obs_cop, "time"))

cat("Observed spatial footprint (Copernicus):\n")
print(summary(cluster_obs$cluster_size))
```

The same statistic applied to the simulated scenarios reveals how well each
model reproduces the observed footprint distribution.

```{r cluster-sim}
cluster_cvine <- hm_cluster_size(scenarios_cvine, threshold = 25)
cluster_dvine <- hm_cluster_size(scenarios_dvine, threshold = 25)
cluster_gauss <- hm_cluster_size(scenarios_gauss, threshold = 25)
cluster_sf    <- hm_cluster_size(scenarios_sf,    threshold = 25)

footprint_table <- data.frame(
  Model  = c("Observed", "C-vine", "D-vine", "Gaussian", "SF (Matérn)"),
  Mean   = round(c(mean(cluster_obs$cluster_size),
                   mean(cluster_cvine$cluster_size),
                   mean(cluster_dvine$cluster_size),
                   mean(cluster_gauss$cluster_size),
                   mean(cluster_sf$cluster_size)), 2),
  Median = round(c(median(cluster_obs$cluster_size),
                   median(cluster_cvine$cluster_size),
                   median(cluster_dvine$cluster_size),
                   median(cluster_gauss$cluster_size),
                   median(cluster_sf$cluster_size)), 2),
  Max    = c(max(cluster_obs$cluster_size),
             max(cluster_cvine$cluster_size),
             max(cluster_dvine$cluster_size),
             max(cluster_gauss$cluster_size),
             max(cluster_sf$cluster_size))
)

print(footprint_table, row.names = FALSE)
```

```{r cluster-plot, fig.height = 5, fig.cap = "Empirical CDF of spatial footprint for observed data and four models."}
all_clusters <- rbind(
  data.frame(model = "Observed",    size = cluster_obs$cluster_size),
  data.frame(model = "C-vine",      size = cluster_cvine$cluster_size),
  data.frame(model = "D-vine",      size = cluster_dvine$cluster_size),
  data.frame(model = "Gaussian",    size = cluster_gauss$cluster_size),
  data.frame(model = "SF (Matérn)", size = cluster_sf$cluster_size)
)

all_clusters$model <- factor(
  all_clusters$model,
  levels = c("Observed", "C-vine", "D-vine", "Gaussian", "SF (Matérn)")
)

ggplot2::ggplot(all_clusters,
                ggplot2::aes(x = size, colour = model)) +
  ggplot2::stat_ecdf(linewidth = 0.9) +
  ggplot2::scale_colour_manual(
    values = c("Observed"    = "black",
               "C-vine"      = "#2166AC",
               "D-vine"      = "#4DAF4A",
               "Gaussian"    = "#FF7F00",
               "SF (Matérn)" = "#E41A1C"),
    name = "Model"
  ) +
  ggplot2::labs(
    title = "Spatial footprint validation — Copernicus",
    x     = "Number of cells \u2265 25 m/s",
    y     = "Empirical CDF"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))
```

<div class="result-box">
**How to read this plot.** A model whose CDF curve lies close to the black
"Observed" curve reproduces the spatial footprint well. A curve to the left
means the model generates events that are too localised; a curve to the right
means events that are too widespread. Use this plot to choose the model that
best captures the dependence structure for your application.
</div>

---

## Summary

```{r summary-table, echo = FALSE}
summary_df <- data.frame(
  Step = paste0("Step ", 1:7),
  Goal = c("Read and standardize a gridded dataset",
           "Isolate extreme-event days",
           "Fit a marginal distribution to each cell",
           "Quantify spatial correlation vs distance",
           "Fit copula or stochastic field model",
           "Generate spatially coherent scenarios",
           "Compare models against observed footprint"),
  `Key functions` = c(
    "`hm_read_netcdf`, `hm_standardize_coords`, `hm_decode_time`",
    "`hm_select_extreme_events`, `hm_to_matrix`, `hm_filter_land_cells`",
    "`hm_fit_marginals`, `hm_pit_transform`",
    "`hm_distance_matrix`, `hm_empirical_kendall`, `hm_empirical_correlogram`",
    "`hm_fit_copula`, `hm_fit_correlation_model`",
    "`hm_simulate_copula`, `hm_simulate_sf_kle`",
    "`hm_cluster_size`"
  ),
  check.names = FALSE
)
knitr::kable(summary_df, align = "clll")
```

---

## References

Clavijo Mesa, M.V., Di Maio, F. & Zio, E. (2026a). A modeling framework
for the inoperability assessment of interdependent critical infrastructures
exposed to spatially distributed natural hazards. *Reliability Engineering &
System Safety*, 275, 112830.

Clavijo Mesa, M.V., Broggi, M., Di Maio, F. & Zio, E. (2026b).
Inoperability assessment of interdependent critical infrastructures exposed
to natural hazards considering climate change. *International Journal of
Disaster Risk Reduction*, 141, 106172.
